Deep Learning Project

### 2024-09-28 project/paper.tex
Update paper and associated references
Corrected typos and clarified details throughout the paper. Modified the methodology and findings sections, improving accuracy statistics and explanations. Updated the bibliography to include a new reference on multi-class image classification using TensorFlow.

\title{Deep Learning Project}
\author{
    Group Members: \\
    Farooq Mahmud
}
\date{September 20, 2024}

@@

This paper aims to replicate the results presented by Szegedy et al., critically assess the impact of the Inception architecture, and test its performance on an alternative dataset. We attempt to reproduce the GoogLeNet model by writing our own Python implementation using the Tensorflow library. Our implementation is evaluated on both a subset of the ImageNet dataset and a secondary dataset, CIFAR-10, to assess its generalization capability. Through this replication and exploration, we seek to validate the core contributions of the Inception model, gain a deeper undertanding of convolutional neural networks, and an appreciation of the associated challenges.

@@

\subsubsection{Inception Architecture}
The Inception architecture was introduced to address two primary challenges in CNNs: improving model accuracy while maintaining computational efficiency. Szegedy et al. proposed a novel structure called the Inception module, which incorporates parallel convolutional filters of different sizes (1x1, 3x3, and 5x5) along with max-pooling. The key advantage of this approach is that it allows the network to capture features at multiple scales simultaneously, enhancing the model's ability to learn complex spatial hierarchies within images.

@@

\subsection{Image Selection}
The Imagenet dataset contains over 1 million images with a size of over 1TB. Processing this amount of data would not be practical given the available coumputing power. In order to make processing possible in a reasonable timeframe, a subset of the Imagenet images was downloaded. The subsets, or \textit{synsets} map to the CIFAR-10 image classes as shown in Table \ref{tab:synsets}.

\begin{table}[ht]
    \centering
    \begin{tabular}{ll}
    \hline
    \textbf{Synset ID} & \textbf{Equivalent CIFAR-10 Class} \\ \hline
    n02691156 & airplane \\ \hline
    n02958343 & automobile \\ \hline
    n01503061 & bird \\ \hline
    n02121808 & cat \\ \hline
    n02419796 & deer \\ \hline
    n02084071 & dog \\ \hline
    n01641577 & frog \\ \hline
    n02374451 & horse \\ \hline
    n04194289 & ship \\ \hline
    n04467665 & truck \\ \hline
    \end{tabular}
    \caption{Imagenet synset to CIFAR-10 class mapping}
    \label{tab:synsets}
\end{table}

The resulting download consists of 14,738 files with a total size of 1.3GB. Scaling the images to 224 x 224 pixels, the dimensions used by Szegedy, et. al. decreased the size to a manageable 137MB.

@@

\subsection{A Simplified GoogLeNet Network}
Before constructing a complex model like GoogLeNet, it is a good idea to use a simple network as a baseline. This network is shown in Figure \ref{fig:cnn-simple}. This network is conceptually similar to GoogLeNet but without the complexity of the Inception modules. At it's core, the GoogLeNet architecture stacks multiple convolutional layers before downsampling through pooling. Stacking multiple layers helps networks capture progressively more complex features, improves generalization, and increases performance while keeping computational cost from getting out of hand. This is a common design pattern in many successful image classification networks\cite{simonyan2014very}.

@@

The validation accuracy and training accuracy track well, indicating a lack of overfitting. Applying data augmentation to the training set mitigates overfitting, especially with the smaller dataset. The augmentation code is shown in Listing \ref{lst:augmentation}

\begin{lstlisting}[caption={Augmentation code}, label={lst:augmentation}]
augmentation = tf.keras.Sequential([
    layers.RandomFlip("horizontal"),
    layers.RandomRotation(0.1),
    layers.RandomZoom(0.2),
])

# Apply the augmentation only on the training dataset
train_dataset = train_dataset.map(lambda x, y: (augmentation(x, training=True), y))
\end{lstlisting}

\begin{figure}[ht]
    \centering
    \includegraphics[scale=0.7]{project/paper_images/googlenet_simple_accuracy.png}
    \caption{Accuracy of the simplified version of the GoogLeNet network}
    \label{fig:googlenet-simple-accuracy}
\end{figure}

\subsection{Replicating the GoogLeNet Network}
The GoogLeNet architecture is explained in tabular form in Table 1 in the Szegedy paper and is reproduced in Figure \ref{fig:googlenet-arch}. The introduction of Inception layers having parallel convolutional filters is the primary difference from typical sequential networks. The replication attempt, when trained for 30 epochs, achieved a maximum accuracy of 56.9\%. This is essentialy the same as the simple network. The upshot is that it only took 8 epochs to arrive at the maximum accuracy as shown in Figure \ref{fig:googlenet-complex-accuracy}. Nevertheless, both models are nowhere near the accuracy of the model in the Szegedy paper. Their model achieved a top-5 error rate of only 6.7\%. While the paper does not mention the accuracy, it is not a leap to assume that the accuracy was well above the accuracy in the replication attempt. The network is graphically depicted in Figure \ref{fig:googlenet-complex}

@@

\begin{figure}[ht]
    \centering
    \includegraphics[scale=0.7]{project/paper_images/googlenet_complex_accuracy.png}
    \caption{Accuracy of the simplified version of the GoogLeNet network}
    \label{fig:googlenet-complex-accuracy}
\end{figure}

@@

\subsubsection{Performance On Unseen Imagenet Images}
100 images from each class was set aside in order to evaluate the performance of the two models on unseen images. The accuracy of the two models is shown in \ref{tab:perf-unseen-imagenet}.

\begin{table}[ht]
    \centering
    \begin{tabular}{ll}
    \hline
    \textbf{Model} & \textbf{Accuracy} \\ \hline
    Simple & 58.1\% \\ \hline
    GoogLeNet Reproduction & ??? \\ \hline
    \end{tabular}
    \caption{Models' performance on unseen ImageNet images}
    \label{tab:perf-unseen-imagenet}
\end{table}

@@

@@

The low accuracy of 21.4\% achieved by the simple model could be for several reasons. First, the models were trained on images having a relatively high resolution compared to CIFAR-10's 32x32 images. When models are trained on larger images, the models learn to identify fine-grained details which can become indistinguishable when downscaled to much smaller dimensions~\cite{simonyan2015vgg}.

Additionally, the convolutional filters in the models are likely too large for the small CIFAR-10 images, causing the network to miss out on important features. In a 32x32 image, for instance, using large filter sizes or pooling layers can result in excessive downsampling, leading to a significant loss of spatial information, which ultimately hampers the model's ability to classify the images accurately.

Finally, CIFAR-10, like ImageNet, contains complex image classes, which are often difficult to distinguish at such a low resolution. The limited information provided by 32x32 images presents additional challenges for models trained on high-resolution images, as they may rely on more detailed textures and edges, which are not as pronounced in low-resolution images~\cite{krizhevsky2012imagenet}.

@@

One of the challenges we encountered was the high computational demand of training, event on the subset of ImageNet images we used. Although GoogLeNet is more efficient than many of its contemporaries, training still required GPU resources. We used data augmentation techniques, such as random cropping, random horizontal flipping, and random zoom to avoid overfitting. We also needed to reduce the learning rate to 0.0001, reduced batch size to 16, and enabled memory growth\cite{tensorflow_memory_growth} so as not to overwhelm the GPU. Prefetching was also implemented to improve performance\cite{tensorflow_data_performance}. The training performance is summarized in Table \ref{tab:training_times}

\begin{table}[ht]
    \centering
    \begin{tabular}{llll}
    \hline
    \textbf{Model} & \textbf{Batch size} & \textbf{Epochs} & \textbf{Training time (minutes)} \\  \hline
    Simple & 16 & 30 & 73 \\ \hline
    Complex & 16 & 30 & 69 \\ \hline
    \end{tabular}
    \caption{Training times}
    \label{tab:training_times}
\end{table}

>>>>

\title{Deep Learning Project}
\author{
    Group Members: \\
    Farooq Mahmud
}
\date{September 28, 2024}

@@

This paper aims to replicate the results presented by Szegedy et al., critically assess the impact of the Inception architecture, and test its performance on an alternative dataset. We attempt to reproduce the GoogLeNet model by writing our own Python implementation using the Tensorflow library. Our implementation is evaluated on both a subset of the ImageNet dataset and a secondary dataset, CIFAR-10, to assess its generalization capability. Through this replication and exploration, we seek to validate the core contributions of the Inception model, gain a deeper understanding of convolutional neural networks, and an appreciation of the associated challenges.

@@

\subsubsection{Inception Architecture}
The Inception architecture was introduced to address the primary challenge in convolutional neural networks- improving model accuracy while maintaining computational efficiency. Szegedy et al. proposed a novel structure called the Inception module, which incorporates parallel convolutional filters of different sizes (1x1, 3x3, and 5x5) along with max-pooling. The key advantage of this approach is that it allows the network to capture features at multiple scales simultaneously, enhancing the model's ability to learn complex spatial hierarchies within images.

@@

\subsection{Image Selection}
The ImageNet dataset contains over 1 million images with a size of over 1TB. Processing this amount of data would not be practical given the available computing power. In order to make processing possible in a reasonable time frame, a subset of the ImageNet images was downloaded. The subsets, or \textit{synsets} map to the CIFAR-10 image classes as shown in Table \ref{tab:synsets}.

\begin{table}[ht]
    \centering
    \begin{tabular}{ll}
    \hline
    \textbf{Synset ID} & \textbf{Equivalent CIFAR-10 Class} \\ \hline
    n02691156 & airplane \\ \hline
    n02958343 & automobile \\ \hline
    n01503061 & bird \\ \hline
    n02121808 & cat \\ \hline
    n02419796 & deer \\ \hline
    n02084071 & dog \\ \hline
    n01641577 & frog \\ \hline
    n02374451 & horse \\ \hline
    n04194289 & ship \\ \hline
    n04467665 & truck \\ \hline
    \end{tabular}
    \caption{Mapping ImageNet synsets to CIFAR-10 classes}
    \label{tab:synsets}
\end{table}

The resulting download consists of 14,738 files with a total size of 1.3GB. Scaling the images to 224 x 224 pixels, the dimensions used by Szegedy, et al. decreased the size to a manageable 137MB.

@@

\subsection{A Simplified GoogLeNet Network}
\label{subsec:simple_model}
Before constructing a complex model like GoogLeNet, it is a good idea to use a simple network as a baseline. This network is shown in Figure \ref{fig:cnn-simple}. This network is conceptually similar to GoogLeNet but without the complexity of the Inception modules. At it's core, the GoogLeNet architecture stacks multiple convolutional layers before down sampling through pooling. Stacking multiple layers helps networks capture progressively more complex features, improves generalization, and increases performance while keeping computational cost from getting out of hand. This is a common design pattern in many successful image classification networks\cite{simonyan2014very}.

@@

The validation accuracy and training accuracy track well, indicating a lack of overfitting. Applying data augmentation to the training set mitigates overfitting, especially with the smaller dataset.
\begin{figure}[ht]
    \centering
    \includegraphics[scale=0.7]{project/paper_images/googlenet_simple_accuracy.png}
    \caption{Accuracy of the simplified version of the GoogLeNet network}
    \label{fig:googlenet-simple-accuracy}
\end{figure}

\subsection{Replicating the GoogLeNet Network}
The GoogLeNet architecture is explained in Table 1 of the Szegedy paper is reproduced in Figure \ref{fig:googlenet-arch}. The introduction of Inception layers having parallel convolutional filters is the primary difference from typical sequential networks. The replication attempt, when trained for 30 epochs, achieved a maximum accuracy of 60.3\%. This is quite improved over the simple model. The other impressive aspect is that it only took 14 epochs to arrive at the maximum accuracy as shown in Figure \ref{fig:googlenet-complex-accuracy}. The conclusion is that the architecture does provide marked improvements over the simpler, more naive model described in \ref{subsec:simple_model}.

Nevertheless, this model is nowhere near the accuracy of the model in the Szegedy paper. Their model achieved a top-5 error rate of only 6.7\%. While the paper does not mention the accuracy, it is not a leap to assume that the accuracy was well above the accuracy in the replication attempt. We are optimistic, however, that our implementation is sound and that with a larger dataset, the accuracy can be improved and approach the accuracy achieved by Szegedy et. al. The network is graphically depicted in Figure \ref{fig:googlenet-complex}

@@

\begin{figure}[ht]
    \centering
    \includegraphics[scale=0.7]{project/paper_images/googlenet_complex_accuracy.png}
    \caption{Accuracy of GoogLeNet network reproduction}
    \label{fig:googlenet-complex-accuracy}
\end{figure}

@@

\subsubsection{Performance On Unseen ImageNet Images}
100 images from each class was set aside in order to evaluate the performance of the two models on unseen images. The accuracy of the two models is shown in \ref{tab:perf-unseen-imagenet}.

\begin{table}[ht]
    \centering
    \begin{tabular}{ll}
    \hline
    \textbf{Model} & \textbf{Accuracy} \\ \hline
    Simple & 58.1\% \\ \hline
    GoogLeNet Reproduction & 66.2\% \\ \hline
    \end{tabular}
    \caption{Models' performance on unseen ImageNet images}
    \label{tab:perf-unseen-imagenet}
\end{table}

@@

The GoogLeNet reproduction resulted in a 66.2\% accuracy on unseen ImageNet images. The higher accuracy of this model can be attributed to the same reasons as the simpler model. In this case, additional complexity did provide a marked increase in accuracy. There is cause for optimism that a result closer to what was achieved by Szegedy, et al. could be matched with a larger training set.

@@

The low accuracy of 21.4\% achieved by the simple model could be for several reasons. First, the models were trained on images having a relatively high resolution compared to CIFAR-10's 32x32 images. When models are trained on larger images, the models learn to identify fine-grained details which can become indistinguishable when down scaled to much smaller dimensions~\cite{simonyan2015vgg}.

Additionally, the convolutional filters in the models are likely too large for the small CIFAR-10 images, causing the network to miss out on important features. In a 32x32 image, for instance, using large filter sizes or pooling layers can result in excessive down sampling, leading to a significant loss of spatial information, which ultimately hampers the model's ability to classify the images accurately.

CIFAR-10, like ImageNet, contains complex image classes, which are often difficult to distinguish at such a low resolution. The limited information provided by 32x32 images presents additional challenges for models trained on high-resolution images, as they may rely on more detailed textures and edges, which are not as pronounced in low-resolution images~\cite{krizhevsky2012imagenet}.

Finally, Tensorflow up scales the images to 224 x 224 which surely results in a tremendous loss of fidelity. The fact that the accuracy is not lower is impressive when considering this loss of fidelity.

@@

One of the challenges we encountered was the high computational demand of training, event on the subset of ImageNet images we used. Although GoogLeNet is more efficient than many of its contemporaries, training still required GPU resources. We used data augmentation techniques, such as random cropping, random horizontal flipping, and random zoom to avoid overfitting\cite{dishasai_multiclass_classification}. We also needed to reduce the learning rate to 0.0001, reduced batch size to 16, and enabled memory growth\cite{tensorflow_memory_growth} so as not to overwhelm the GPU. Prefetching was also implemented to improve performance\cite{tensorflow_data_performance}. The training performance is summarized in Table \ref{tab:training_times}. From the training times, it should be readily apparent why a subset of the ImageNet images were chosen and that training the whole dataset would take an inordinate amount of time with the compute resources we had access to.

\begin{table}[ht]
    \centering
    \begin{tabular}{llll}
    \hline
    \textbf{Model} & \textbf{Batch size} & \textbf{Epochs} & \textbf{Training time (minutes)} \\  \hline
    Simple & 16 & 30 & 58 \\ \hline
    Complex & 16 & 30 & 64 \\ \hline
    \end{tabular}
    \caption{Training times}
    \label{tab:training_times}
\end{table}
